---
title: "TESSERA Quickstart: Coral Bleaching Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TESSERA Quickstart: Coral Bleaching Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_coral <- tryCatch({
  data(coral_example, package = "tessera")
  TRUE
}, error = function(e) FALSE)
```

## Overview

TESSERA (Topology-Enhanced Similarity and Structure for Entity Regime
Analysis) detects regime boundaries in complex systems by analyzing how
similarity network topology shifts over time. This vignette demonstrates the
full Phase 1 pipeline using coral bleaching data.

## The Pipeline

TESSERA's workflow has three analytical stages (Phase 1):

1. **Characterize** — bin entities temporally, prepare features
2. **Discover** — compute similarity, build networks, detect communities
3. **Quantify** — measure regime divergence across bins

```{r library}
library(tessera)
```

## Load Data

The package includes a curated subset of the Global Coral Bleaching Database
(van Woesik et al. 2022), stratified across temporal quartiles and bleaching
outcomes.

```{r load-data}
if (has_coral) {
  data(coral_example)
  df <- coral_example
  time_col <- "year"
  cat("Coral bleaching subset:", nrow(df), "records,",
      ncol(df), "features\n")
  cat("Year range:", range(df[[time_col]]), "\n")
} else {
  # Fallback: synthetic data with ecological structure
  set.seed(42)
  n <- 500
  df <- data.frame(
    year = sample(1990:2020, n, replace = TRUE),
    sst = rnorm(n, 27, 2),
    depth = runif(n, 1, 40),
    turbidity = rlnorm(n, 1, 0.5),
    coral_cover = rbeta(n, 2, 5)
  )
  time_col <- "year"
  cat("Using synthetic data (coral_example not available)\n")
}
```

## Quick Start

The fastest path is the `tessera()` wrapper with a domain preset:

```{r quick}
result <- tessera(df, time_col = time_col)
result
```

## Step-by-Step Pipeline

For full control, use the pipeline functions individually. Each returns an
inspectable S3 object.

### Step 1: Characterize

```{r characterize}
char <- tessera_characterize(df, time_col = time_col, n_bins = 4)
char
```

Key decisions at this stage:

- **`bin_method = "quantile"`** (default): ensures balanced entity counts.
  Critical for datasets like coral bleaching where early decades have sparse
  records.
- **`scale = TRUE`** (default): standardizes features within each bin so that
  variables on different scales contribute equally to similarity.
- **`min_entities = 30`**: bins with fewer entities are merged automatically.

```{r char-summary}
summary(char)
```

### Step 2a: Compute Similarity

```{r similarity}
sim <- tessera_similarity(char,
                          measures = c("bray_curtis", "simpson"),
                          k = 15)
sim
```

The ecology preset uses Bray-Curtis (from `vegan`, the standard ecological
dissimilarity) and Simpson (quantile-bin overlap — TESSERA's signature
measure). The sparse kNN representation keeps only the top-k neighbors per
entity, which is essential for large datasets.

### Step 2b: Build Networks

```{r network}
net <- tessera_network(sim, method = "knn", k = 12)
net
```

### Step 3a: Detect Communities

```{r communities}
com <- tessera_communities(net, algorithm = "louvain", resolution = 1.0)
com
```

Communities represent groups of entities with similar feature profiles.
In ecological terms, these correspond to reef assemblages or environmental
regimes.

### Step 3b: Quantify Regime Divergence

```{r divergence}
div <- tessera_divergence(com, method = "nmi", reference = "adjacent")
div
```

Regime divergence (RD) measures how community structure changes between
adjacent temporal bins. High RD indicates a regime boundary — an "ecotone"
in biogeographical terminology.

### Visualization

```{r plot-div, fig.cap = "Regime divergence across temporal bins"}
plot(div)
```

## Interpreting Results

TESSERA identifies three failure geometries that manifest as distinct RD
patterns:

- **Boundary-concentrated**: High RD in one or two transitions, low
  elsewhere. Indicates a clear regime shift (e.g., a bleaching threshold
  crossing).
- **Isolated cluster**: A sudden spike at one bin transition. Suggests an
  anomalous period rather than a sustained shift.
- **Diffuse**: Gradually increasing RD across all transitions. Indicates
  continuous environmental drift.

## Using the Full Coral Database

The bundled `coral_example` is a 500-record subset. To analyze the full
Global Coral Bleaching Database (31,199 records):

```{r real-data, eval = FALSE}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "coral_bleaching.sqlite")
coral_full <- dbReadTable(con, "bleaching_records")
dbDisconnect(con)

result <- tessera(coral_full,
                  time_col = "year",
                  preset = "ecology")
```

## Next Steps

- Use `plot(result, which = "all")` for a diagnostic panel
- Access intermediate objects: `result$communities$by_bin$bin_1$membership`
- Compare similarity measures: run with `measures = "all"`
- Adjust resolution to find finer or coarser community structure
